---
title: "QTLExperiment and multistateQTL: Simulation examples"
author: "C.B. Del Azodi"
date: "`r Sys.Date()`"
output: html_document
---

## Set up environment

```{r knitr-options, echo = FALSE}
knitr::opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE, 
                      fig.height = 6, fig.width = 8)

```

```{r load-packages}
suppressPackageStartupMessages({
  library(ggpubr)
  library(dplyr)
  library(ComplexHeatmap)
  library(cowplot)
  library(QTLExperiment)
  devtools::load_all("/mnt/mcfiles/cazodi/Software/multiStateQTL")
})

set.seed(42)
rerun <- FALSE
plot_theme <- theme_cowplot(8)

save <- TRUE
save.loc <- paste0("output/00_Figures/", Sys.Date(), "_")
params <- readRDS("output/gtex_estimated_params.rds")

# Simulation parameters
n_tests <- 500
n_states <- 6
thresh <- 0.1
second_thresh <- 0.2
factor <- 0.75
sharing <- c(0, 1)
```


# Simulation examples

### Simple simulation

Using the parameters estimated from the GTEx data, here we simulate a data set with 80% of QTL effects shared across all states and the remaining 20% of QTL effects unique to a single state. 

```{r global-QTL-simulation}
sim_simple <- qtleSimulate(params=params, ntests=n_tests, nstates=n_states, 
                           global=0.8, unique=0.2) 

sim_simple <- callSignificance(sim_simple, assay="lfsrs", 
                               thresh=thresh, second.thresh = second_thresh)
sim_simple <- runPairwiseSharing(sim_simple, assay="significant", factor=factor)
p_simple <- plotPairwiseSharing(sim_simple, col_range=sharing,
                                cell_annotate = TRUE)
draw(p_simple)

if(save){
  tmp <- paste0(save.loc, "sim_simple_pairwise.pdf")
  message("Saving to: ", tmp)
  pdf(tmp, height=5, width=5.5)
  draw(p_simple)
  while (!is.null(dev.list()))  dev.off()
}
```


### Variable power simulation

Simulating a data set again with 80% of QTL effects shared across all states and the remaining 20% of QTL effects unique to a single state. However, in this simulation, the betas for each state are sampled using state-specific standard deviations. This situation could occur if you have less data available for some states and therefore expect more variance in the estimated QTL effect sizes. Here we simulate S1-S4 with sd=0.1 (default), S5 with sd=1, and S6 with sd=2. 

```{r variable-sd-QTL-simulation}
sim_sd <- qtleSimulate(params=params, ntests=n_tests, nstates=n_states, 
                       global=0.8, unique=0.2, 
                       beta_sd=c(0.1, 0.1, 0.1, 0.1, 1, 2))

sim_sd <- callSignificance(sim_sd, assay="lfsrs", 
                               thresh=thresh, second.thresh = second_thresh)
sim_sd <- runPairwiseSharing(sim_sd, assay="significant", factor=factor)
p_sd <- plotPairwiseSharing(sim_sd, col_range=sharing,
                                cell_annotate = TRUE)
draw(p_sd)

if(save){
  tmp <- paste0(save.loc, "sim_sd_pairwise.pdf")
  message("Saving to: ", tmp)
  pdf(tmp, height=5, width=5.5)
  draw(p_sd)
  while (!is.null(dev.list()))  dev.off()
}
```

### Multi-state simulation

Simulating a dataset where states are clustered in groups, with 50% of QTL assigned randomly to only one cluster and 50% shared across all states. Here we simulate two clusters, but k can be set to any number smaller than the number of states.

```{r multistate-QTL-simulation}
sim_multi <- qtleSimulate(params=params, ntests=n_tests, nstates=n_states, 
                          global=0.5, multi=0.5, k=c(1, 1, 1, 1, 2, 2))

sim_multi <- callSignificance(sim_multi, assay="lfsrs", 
                               thresh=thresh, second.thresh = second_thresh)
sim_multi <- runPairwiseSharing(sim_multi, assay="significant", factor=factor)
p_multi1 <- plotPairwiseSharing(sim_multi, col_range=sharing,
                                cell_annotate = TRUE,
                                annotate_rows = "multistateGroup")
draw(p_multi1)


if(save){
  tmp <- paste0(save.loc, "sim_multistate_pairwise.pdf")
  message("Saving to: ", tmp)
  pdf(tmp, height=5, width=5.5)
  draw(p_multi1)
  while (!is.null(dev.list()))  dev.off()
}
```

To get more detailed insight into the patterns of QTL sharing across states, we can also visualize sharing with UpSet plots. 

```{r multi-state-upset-plot, fig.height=5, fig.width=6, fig.cap="Upset plot of significant QTL tests for the multi-state simulation."}
p_upset <- plotUpSet(sim_multi, annotate_by = c("multistateGroup", "nSignificant"), 
                     set_order = NULL)
draw(p_upset)

if(save){
  tmp <- paste0(save.loc, "multistate_upset.pdf")
  message("Saving to: ", tmp)
  pdf(tmp, height=5, width=6)
  draw(p_upset)
  while (!is.null(dev.list()))  dev.off()
}
```


# QTL test level analysis

## QTL Classification 

The `runTestMetrics` function outputs helpful information about each QTL test, including the number of states where the test was significant, the standard deviation of the betas across states, and the classification of that QTL. Here we compare the simulate QTL types (columns: global, multistate, unique) to the QTL classifications based on our analysis pipeline (rows: global, multistate, unique, not significant).

```{r multi-state-simulation-types}
sim_multi <- runTestMetrics(sim_multi)
sim_stats <- table(rowData(sim_multi)$qtl_type_simple, 
                   rowData(sim_multi)$QTL)
sim_stats[c("global", "multistate", "unique"), ]
```

## QTL visualizations

The `plotQTLClusters` function automates plotting a heatmap of betas or any other user defined assay for tests and states. Here we show simulated betas from the multi-state simulation and annotate the plot with helpful information from the colData and rowData.

```{r plot-multistate-betas, fig.height=8, fig.width=6, fig.cap=""}
rowData(sim_multi)$mean_beta_abs <- abs(rowData(sim_multi)$mean_beta)
sim_multi_subset <- subset(sim_multi, QTL == "multistate")

p_ms_clusters <- plotQTLClusters(sim_multi_subset,
                                 row_km = 5, 
                                 annotate_states = c("multistateGroup"),
                                 annotate_tests = c("mean_beta", "nSignificant"))

draw(p_ms_clusters)

if(save){
  tmp <- paste0(save.loc, "multistate_clusters.pdf")
  message("Saving to: ", tmp)
  pdf(tmp, height=7, width=6)
  draw(p_ms_clusters)
  while (!is.null(dev.list()))  dev.off()
}
```

